*Symbiodinium* communities in reef corals of St. John, U.S.V.I. (2012)

### 2017-03-26 manuscript/manuscript.Rmd
submission

---
title: Describing *Symbiodinium* metacommunities using high-throughput sequencing
  of ITS2
author:
- affiliation: University of Hawaii
  email: ross.cunning@gmail.com
  footnote: Corresponding Author
  name: Ross Cunning
- affiliation: University of Hawaii
  name: Ruth D. Gates
- affiliation: California State University Northridge
  name: Peter J. Edmunds
bibliography: library.bib
output:
  word_document: default
  pdf_document: default
  html_document:
    fig_caption: yes
subtitle: A case study from St. John, U.S. Virgin Islands
address:
- address: Hawaii Institute of Marine Biology, Kaneohe, HI 96744, USA
  code: University of Hawaii
- address: Department of Biology, Northridge, CA 91330, USA
  code: California State University Northridge
---
```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = normalizePath(".."))
knitr::opts_chunk$set(echo = F, warning = F, message = F, cache = T)
options(scipen = 3, digits = 9)
```

@@

### Abstract

  Symbiotic microalgae (*Symbiodinium* spp.) strongly influence the performance and stress-tolerance of their coral hosts, making the analysis of *Symbiodinium* communities in corals (and metacommunities on reefs) advantageous for many aspects of coral reef research. High-throughput sequencing of ITS2 nrDNA offers unprecedented scale in describing these communities, yet high intragenomic variability at this locus complicates the resolution of biologically meaningful diversity. Here, we demonstrate that generating operational taxonomic units by clustering ITS2 sequences at 97% similarity within, but not across, samples collapses sequence diversity that is more likely to be intragenomic, while preserving diversity that is more likely interspecific. We utilize this 'within-sample clustering' to analyze *Symbiodinium* from ten host taxa on shallow reefs on the north and south shores of St. John, US Virgin Islands. While *Symbiodinium* communities did not differ between shores, metacommunity network analysis of host-symbiont associations revealed *Symbiodinium* lineages occupying 'dominant' and 'background' niches, and coral hosts that are more 'flexible' or 'specific' in their associations with *Symbiodinium*. These methods shed new light on important questions in coral symbiosis ecology, and demonstrate how application-specific bioinformatic pipelines can improve the analysis of metabarcoding data in microbial metacommunity studies.

### Introduction

  The composition of symbiotic algal communities (*Symbiodinium* spp.) in reef-building scleractinian and milleporine corals plays a major role in their biology and ecology, as the identity and functional performance of the symbionts influences emergent properties of the holobiont, including its photobiology, energetics, growth rates, and susceptibility to stress [@Rowan:2004p3383; @Cantin:2009p4376; @Jones:2011dt]. Even slight differences in the relative abundance of different symbionts may have important functional consequences for the host [@Loram:2007p3688; @Cunning:2015ja]. Moreover, variation in these communities among individuals within a host species, and within individuals over time, is an important trait modulating sensitivity of corals to environmental stress [@Putnam:2012wv] and their ability to respond in beneficial ways to changing environmental conditions by 'shuffling' or 'switching' symbionts [@Baker:2003p200]. Therefore, understanding the ecological performance of corals will greatly benefit from accurate and comprehensive descriptions of their *Symbiodinium* communities.

  *Symbiodinium* identity is primarily described using genetic sequence data, including both chloroplast (cp23S and psbA^ncr^) and nuclear markers (18S, 28S, ITS1, and ITS2) [@LaJeunesse:2001p64; @Santos:2003kt; @Pochon:2012we]. Together, these markers have been used to identify nine major 'clades' within the genus *Symbiodinium* (clades A through I [@Pochon:2010p8784]), which have been further divided into 'types' based on the marker used to identify them. *Symbiodinium* species are beginning to be described based on combinations of markers, including microsatellites to establish reproductive isolation (i.e., to satisfy the biological species concept) [@LaJeunesse:2014io]. However, ecological surveys of *Symbiodinium* diversity must still rely on commonly-used marker genes, such as ITS2. Consequently, high-throughput sequencing of ITS2 is being utilized to characterize *Symbiodinium* communities with unprecedented resolution [e.g., @Arif:2014gk; @Quigley:2014dv; @Thomas:2014da; @Edmunds:2014jc; @Cunning:2015hb; @Ziegler:2017cs]. However, such analyses may create datasets consisting of millions of sequence reads and hundreds of thousands of distinct sequence variants [@Ziegler:2017cs], which places great importance on the ways in which ITS2 sequence diversity is analyzed and interpreted in relation to biological diversity.

@@

### Materials and Methods

#### Sample collection and environmental conditions

@@

#### ITS2 sequencing and bioinformatics

  Coral tissue samples were preserved in ~500 µL Guanidinium buffer (50% w/v guanidinium isothiocyanate; 50 mM Tris pH 7.6; 10 μM EDTA; 4.2% w/v sarkosyl; 2.1% v/v β-mercaptoethanol) and shipped to the Hawaii Institute of Marine Biology (HIMB). Genomic DNA was extracted from each coral tissue sample following a guanidinium-based extraction protocol [@Cunning:2015hb], and sent to Research and Testing Laboratory (Lubbock, TX) for sequencing of ITS2 amplicons ('itsD' and 'its2rev2' primers from @Stat:2009p7366) on the Illumina MiSeq platform with 2x300 paired-end read chemistry.

  Paired reads from each sample (provided in .fastq format by Research and Testing Laboratory) were merged using illumina-utils software [@Eren:2013hh] with an enforced Q30-check and an overlap ≥150 bp with ≤3 mismatches required to retain a sequence. Chimeric sequences were removed using usearch61 [@Edgar:2010cv] implemented in QIIME [@Caporaso:2010jf]. Primers were trimmed using cutadapt [@Martin:2011eu] allowing 3 mismatches, and only sequences with both forward and reverse primer matches and length ≥250 bp after trimming were retained. Subsequently, three different clustering approaches, each based on the uclust algorithm [@Edgar:2010cv] and implemented in QIIME, were used to group ITS2 sequences into operational taxonomic units (OTUs): 1) clustering at 100% identity, 2) clustering at 97% identity across samples (i.e., sequences from all samples clustered together), and 3) clustering at 97% identity within samples (i.e., sequences from each sample clustered independently). For each 97% cluster, the most abundant sequence variant was chosen as the representative sequence, and within-sample clusters were merged across samples if their representative sequences were 100% identical. After removing singleton clusters, representative sequences for each OTU were assigned taxonomy by searching a custom reference database of *Symbiodinium* ITS2 sequences using the Needleman-Wunsch global alignment algorithm implemented in the Biostrings package [@BiostringsStringo:Fbauxymn] in R [@RALanguageandEn:2014wf]. Each OTU was then assigned a name corresponding to the reference sequence(s) with the highest alignment score; if the match was <100%, the assignment was given a unique superscript. If the match was <90%, the sequence was blasted to NCBI and omitted if the top hit did not contain "*Symbiodinium*".

  The reference database comprised `r system("grep -c '>' data/ITS2db_trimmed_derep.fasta")` *Symbiodinium* ITS2 sequences downloaded directly from NCBI. These sequences included those used in previous reference databases [@Cunning:2015hb] supplemented with additional sequences of *Symbiodinium* from Caribbean corals [@Finney:2010p8324; @Green:2014ev]. Reference sequences were separated and aligned by clade using muscle [@Edgar:2004bo], and trimmed to equal length using the o-smart-trim command from oligotyping software [@Eren:2013fp], before being reconcatenated into a single fasta file. The bioinformatic pipeline used here was implemented using a series of bash scripts and a Makefile, which can be found in the data archive accompanying this paper and executed to fully reproduce the present analysis.

#### *Symbiodinium* data analysis

@@

  Network analysis and visualization was performed in R using the igraph package [@Csardi:2006vh]. Networks were created featuring 'dominant' (>50% relative abundance) and 'abundant' (>1% relative abundance) OTUs, with weighted edges proportional to the number of individuals within a species in which a symbiont OTU occurred. A 'background' symbionts network was also created with unweighted edges defined based on the presence of a symbiont at <1% relative abundance in at least one individual within the species. To simplify visualization of the background symbiont network, clade D OTUs were merged into a single node under the assumption that clade D in the Caribbean comprises a single species (*Symbiodinium trenchii*; [@LaJeunesse:2014io]), and symbiont nodes connected to ≤ 2 coral species were removed from the network. All network layouts were constructed based on the Fruchterman-Reingold algorithm [@Fruchterman:1991ko].

@@

### Results

#### Sample collection and environmental data

@@

#Wave exposure
we <- list(N=3.234787362,	S=4.025022708)
barplot(c(we$N, we$S), horiz=T, xlim=c(0,5), ylab="", xlab="Wave exposure (ln J m-3)")
text(par("usr")[1], c(1,2), c("S","N"), pos=2, offset=1, xpd=NA)
points(rep(par("usr")[1]-strwidth("c.m"),2), c(1,2), pch=1, cex=2.5, xpd=NA)

@@

#### Comparison of clustering approaches

@@

knitr::kable(cbind(stats97.f, stats97bs.f, stats100.f)[c(2, 3, 6, 8), ], align="c",
             caption = "Table 1. Summary statistics for each clustering approach.")
```

@@

```{r Fig2, fig.width=6, fig.height=4, fig.align="center", fig.cap="Figure 2. Comparison of dominant OTUs assigned by 100%, 97% across-sample, and 97% within-sample clustering approaches. A) The composition of unique sequence variants (i.e., 100% OTUs) in each sample, sorted by relative abundance, and filled by unique colors corresponding to unique sequence variants. The dominant OTUs within each sample assigned by B) 97% across-sample clustering and C) 97% within-sample clustering are shown by rectangles below each bar, with fill colors matching the unique sequence variants presented in (A) to indicate the representative sequence of the assigned OTU. Colored rectangles that span multiple bars indicate that the corresponding samples were assigned the same dominant OTU. Rectangles are annotated with the taxonomic assignment of the OTU (see Methods)."}
set.seed(42432) 
par(mfrow=c(1,1))
otubarplot2(samples=c("64","72","29","45","73","59","47","26","48","66","63"))
```

#### *Symbiodinium* community composition

@@

```{r Fig3, fig.height=6, fig.width=4, fig.align="center", fig.cap="Figure 3. *Symbiodinium* community composition of each sample. Samples are plotted as horizontal bars, sorted by species and shore of collection (north vs. south). Individual segments of each bar represent 97% within-sample OTUs, colored by *Symbiodinium* clade identity."}
#png(file="crap.png", width = 4, height=6, units="in", res=300)
par(mfrow=c(1,1), mar=c(2, 1.5, 2, 5), lwd=0.1, cex=1)
# Plot composition of 97% within-sample OTUs colored by clade
composition(phy97bs.f.p, col=taxcolors[factor(data.frame(tax_table(phy97bs.f.p))[order(data.frame(tax_table(phy97bs.f.p))$Subtype), ]$Clade, levels=c("A","B","C","D","F","G"))], legend=T)
#dev.off()
```

#### Differences in *Symbiodinium* between shores

@@

```{r Table2, cache=T } 
set.seed(43789)  
shorestats97bs <- perms(phy97bs.f.p, group="Species", trt="Shore")
shorestats97bs$Species <- str_replace_all(
  shorestats97bs$Species,
  c("alcicornis"="*Millepora alcicornis*",
    "annularis"="*Orbicella annularis*",
    "astreoides"="*Porites astreoides*",
    "cavernosa"="*Montastraea cavernosa*",
    "cylindrus"="*Dendrogyra cylindrus*",
    "fragum"="*Favia fragum*",
    "furcata"="*Porites furcata*",
    "radians"="*Siderastrea radians*",
    "siderea"="*Siderastrea siderea*",
    "strigosa"="*Pseudodiplora strigosa*")
)
knitr::kable(shorestats97bs, digits=3, row.names=F, 
             caption="Table 2. Mean overall, within-shore ('within'), and between-shore ('between') Bray-Curtis dissimilarities of the *Symbiodinium* communities in each host species, and PERMANOVA tests (partial R^2^ and p-values) for a difference between shores. Between-shore tests were not possible for *O. annularis* and *P. astreoides* since they were only sampled from one shore.")

@@

#### Network analysis of *Symbiodinium* metacommunity
```{r networks,  cache=F, include=FALSE}
# Create dominant symbionts network
set.seed(12374)
domnet <- sppnet(phy=phy97bs.f.p, plot=F,
                 fun=function(x) length(x[x>0.5])/length(x), 
                 layout=layout.fruchterman.reingold)
# Calculate number of dominant symbionts connected to each species
dom.Nedges <- sapply(levels(data.frame(sample_data(phy97bs.f.p))$Species),
                     FUN=function(x) length(E(domnet)[from(x)]))
dom.Nedges <- sort(dom.Nedges, decreasing=T)

@@

```{r Fig4, fig.width=5, fig.height=5, fig.align="center", fig.retina=1, out.width=5, fig.cap="<br>Figure 4. Network analysis of abundant *Symbiodinium* taxa in each coral species. *Symbiodinium* OTUs (circular nodes) are connected to each coral species (square nodes) in which they ever occurred at >1% relative abundance within an individual; thickness of edges (i.e., links between coral and symbiont nodes) is relative to the proportion of individuals within the coral species in which the *Symbiodinium* OTU was present at >1%. Symbiont nodes are colored by clade identity, and sized proportionally to the number of coral species in which they were found."}

@@

```{r Fig5, fig.width=7.20472, fig.height=3.60236, fig.align="center", fig.cap="Figure 5. Network analysis of dominant (A) and background (B) *Symbiodinium* taxa in each coral species (see Materials and Methods for details on network construction). Symbiont nodes are colored by clade identity, sized proportionally to the number of coral species to which they are connected."}

@@

#### Symbiotic flexibility of host species

@@

```{r Fig6, cache=F, fig.width=3.5, fig.height=3.5, fig.cap="Figure 6. Symbiotic flexibility in coral species quantified as multivariate dispersion of *Symbiodinium* community composition (mean distance to centroid ± SE. Host taxa that do not share a letter are significantly different (p < 0.05)."}
betad97bs <- betad(phy97bs.f.t, group="Species")

@@

### Discussion

#### Bioinformatic analysis of ITS2 diversity

@@

#### *Symbiodinium* metacommunity ecology in St. John

@@

  In addition to hosting apparent generalist ITS2 lineages, most coral species sampled in St. John (except *M. cavernosa* and *P. astreoides*) hosted abundant *Symbiodinium* OTUs that were not abundant in any other corals, and are thus apparently more host-specific. Unique OTUs in the B1-radiation were abundant in *M. alcicornis* and *D. cylindrus*, while others from both the B1- and B19-radiations [see @LaJeunesse:2004p122] were abundant in *F. fragum*, *S. radians*, and *P. strigosa*. Several clade C OTUs similarly were only abundant in one species: C1144 in *P. strigosa*, C7 in *O. annularis*, C1 in *S. siderea*, C1c/C45 and C45a in *P. furcata*, and C46 in *S. radians*. Interestingly, *S. radians* was the only species in which none of the abundant symbionts occurred in any other host, which may reflect the unique ecology of *S. radians* as the only study species that typically forms small, encrusting colonies as adults (XXX).

@@

#### Conclusions

@@

### Acknowledgements

@@

### References

>>>>

---
title: 'Describing *Symbiodinium* metacommunities using high-throughput sequencingof ITS2: A case study from St. John, U.S. Virgin Islands'
author:
- affiliation: HIMB
  email: ross.cunning@gmail.com
  footnote: Corresponding Author
  name: Ross Cunning
- affiliation: HIMB
  name: Ruth D. Gates
- affiliation: CSUN
  name: Peter J. Edmunds
address:
- address: Hawaii Institute of Marine Biology, School of Ocean and Earth Science and
    Technology, University of Hawaii at Manoa, USA
  code: HIMB
- address: Department of Biology, California State University, Northridge, CA, USA
  code: CSUN
output: word_document
#output:
#  rticles::elsevier_article:
#    fig_caption: yes
csl: the-isme-journal.csl
bibliography: library.bib

abstract: |
  Symbiotic microalgae (*Symbiodinium* spp.) strongly influence the performance and stress-tolerance of their coral hosts, making the analysis of *Symbiodinium* communities in corals (and metacommunities on reefs) advantageous for many aspects of coral reef research. High-throughput sequencing of ITS2 nrDNA offers unprecedented scale in describing these communities, yet high intragenomic variability at this locus complicates the resolution of biologically meaningful diversity. Here, we demonstrate that generating operational taxonomic units by clustering ITS2 sequences at 97% similarity within, but not across, samples collapses sequence diversity that is more likely to be intragenomic, while preserving diversity that is more likely interspecific. We utilize this 'within-sample clustering' to analyze *Symbiodinium* from ten host taxa on shallow reefs on the north and south shores of St. John, US Virgin Islands. While *Symbiodinium* communities did not differ between shores, metacommunity network analysis of host-symbiont associations revealed *Symbiodinium* lineages occupying 'dominant' and 'background' niches, and coral hosts that are more 'flexible' or 'specific' in their associations with *Symbiodinium*. These methods shed new light on important questions in coral symbiosis ecology, and demonstrate how application-specific bioinformatic pipelines can improve the analysis of metabarcoding data in microbial metacommunity studies.
---

**Conflict of interest:** The authors declare no conflict of interest.  
**Subject category:** Microbial population and community ecology

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = normalizePath(".."))
knitr::opts_chunk$set(echo = F, warning = F, message = F, cache = T)
options(scipen = 3, digits = 9)
```

@@

# Introduction

  The composition of symbiotic algal communities (*Symbiodinium* spp.) in reef-building scleractinian and milleporine corals plays a major role in their biology and ecology, as identity and the functional performance of symbionts influences emergent properties of the holobiont, including its photobiology, energetics, growth rates, and susceptibility to stress [@Rowan:2004p3383; @Cantin:2009p4376; @Jones:2011dt]. Even slight differences in the relative abundance of different symbionts may have important functional consequences for the host [@Loram:2007p3688; @Cunning:2015ja]. Moreover, variation in these communities among individuals within a host species, and within individuals over time, is an important trait modulating sensitivity of corals to environmental stress [@Putnam:2012wv] and their ability to respond in beneficial ways to changing environmental conditions by 'shuffling' or 'switching' symbionts [@Baker:2003p200]. Therefore, an accurate and comprehensive description of *Symbiodinium* communities informs understanding of the ecological performance of corals.

  *Symbiodinium* identity is primarily described using genetic sequences of chloroplast (cp23S and psbA^ncr^) and nuclear markers (18S, 28S, ITS1, and ITS2) [@LaJeunesse:2001p64; @Santos:2003kt; @Pochon:2012we]. Together, these markers have been used to identify nine major 'clades' within the genus *Symbiodinium* (clades A through I [@Pochon:2010p8784]), which have been further divided into 'types' based on the marker used to identify them. *Symbiodinium* species are beginning to be described based on combinations of markers, including microsatellites to establish reproductive isolation (i.e., to satisfy the biological species concept) [@LaJeunesse:2014io]. However, ecological surveys of *Symbiodinium* diversity still generally rely on commonly-used marker genes, such as ITS2. Consequently, high-throughput sequencing of ITS2 is being utilized to characterize *Symbiodinium* communities with unprecedented resolution [e.g., @Arif:2014gk; @Quigley:2014dv; @Thomas:2014da; @Edmunds:2014jc; @Cunning:2015hb; @Ziegler:2017cs]. However, such analyses often create datasets consisting of millions of sequence reads and hundreds of thousands of distinct sequence variants [@Ziegler:2017cs], which places great importance on the ways in which ITS2 sequence diversity is analyzed and interpreted in relation to biological diversity.

@@

# Materials and Methods

## Sample collection and environmental conditions

@@

## ITS2 sequencing and bioinformatics

  Coral tissue samples were preserved in ~500 µL Guanidinium buffer (50% w/v guanidinium isothiocyanate; 50 mM Tris pH 7.6; 10 $\mu$M EDTA; 4.2% w/v sarkosyl; 2.1% v/v $\beta$-mercaptoethanol) and shipped to the Hawaii Institute of Marine Biology (HIMB). Genomic DNA was extracted from each coral tissue sample following a guanidinium-based extraction protocol [@Cunning:2015hb], and sent to Research and Testing Laboratory (Lubbock, TX) for sequencing of ITS2 amplicons ('itsD' and 'its2rev2' primers from @Stat:2009p7366) on the Illumina MiSeq platform with 2x300 paired-end read chemistry.

  Paired reads from each sample (provided in .fastq format by Research and Testing Laboratory) were merged using illumina-utils software [@Eren:2013hh] with an enforced Q30-check and an overlap $\geq$ 150 bp with $\leq$ 3 mismatches required to retain a sequence. Chimeric sequences were removed using usearch61 [@Edgar:2010cv] implemented in QIIME [@Caporaso:2010jf]. Primers were trimmed using cutadapt [@Martin:2011eu] allowing 3 mismatches, and only sequences with both forward and reverse primer matches and length $\geq$ 250 bp after trimming were retained. Subsequently, three different clustering approaches, each based on the uclust algorithm [@Edgar:2010cv] and implemented in QIIME, were used to group ITS2 sequences into operational taxonomic units (OTUs): 1) clustering at 100% identity, 2) clustering at 97% identity across samples (i.e., sequences from all samples clustered together), and 3) clustering at 97% identity within samples (i.e., sequences from each sample clustered independently). For each 97% cluster, the most abundant sequence variant was chosen as the representative sequence, and within-sample clusters were merged across samples if their representative sequences were 100% identical. After removing singleton clusters, representative sequences for each OTU were assigned taxonomy by searching a custom reference database of *Symbiodinium* ITS2 sequences using the Needleman-Wunsch global alignment algorithm implemented in the Biostrings package [@BiostringsStringo:Fbauxymn] in R [@RALanguageandEn:2014wf]. Each OTU was then assigned a name corresponding to the reference sequence(s) with the highest alignment score; if the match was <100%, the assignment was given a unique superscript. If the match was <90%, the sequence was blasted to NCBI and omitted if the top hit did not contain "*Symbiodinium*".

  The reference database comprised `r system("grep -c '>' data/ITS2db_trimmed_derep.fasta")` *Symbiodinium* ITS2 sequences downloaded directly from NCBI. These sequences included those used in previous reference databases [@Cunning:2015hb] supplemented with additional sequences of *Symbiodinium* from Caribbean corals [@Finney:2010p8324; @Green:2014ev]. Reference sequences were separated and aligned by clade using muscle [@Edgar:2004bo], and trimmed to equal length using the o-smart-trim command from oligotyping software [@Eren:2013fp], before being reconcatenated into a single fasta file. The bioinformatic pipeline used here was implemented using a series of bash scripts which can be found in the data archive accompanying this paper (http://github.com/jrcunning/STJ2012). These scripts are provided along with all raw data and a Makefile which can be executed to fully reproduce the present analysis.

## *Symbiodinium* data analysis

@@

  Network analysis and visualization was performed in R using the igraph package [@Csardi:2006vh]. Networks were created featuring 'dominant' (>50% relative abundance) and 'abundant' (>1% relative abundance) OTUs, with weighted edges proportional to the number of individuals within a species in which a symbiont OTU occurred. A 'background' symbionts network was also created with unweighted edges defined based on the presence of a symbiont at <1% relative abundance in at least one individual within the species. To simplify visualization of the background symbiont network, clade D OTUs were merged into a single node under the assumption that clade D in the Caribbean comprises a single species (*Symbiodinium trenchii*; [@LaJeunesse:2014io]), and symbiont nodes connected to $\leq$ 2 coral species were removed from the network. All network layouts were constructed based on the Fruchterman-Reingold algorithm [@Fruchterman:1991ko].

@@

# Results

## Sample collection and environmental data

@@

#Wave exposure
we <- list(N=3.234787362,	S=4.025022708)
barplot(c(we$S, we$N), horiz=T, xlim=c(0,5), ylab="", xlab="Wave exposure (ln J m-3)")
text(par("usr")[1], c(1,2), c("S","N"), pos=2, offset=1, xpd=NA)
points(rep(par("usr")[1]-strwidth("c.m"),2), c(1,2), pch=1, cex=2.5, xpd=NA)

@@

## Comparison of clustering approaches

@@

knitr::kable(cbind(stats97.f, stats97bs.f, stats100.f)[c(2, 3, 6, 8), ],
             caption = "Summary statistics for each clustering approach.")
```

@@

```{r Fig2, fig.width=6, fig.height=4, fig.cap="Figure 2. Comparison of dominant OTUs assigned by 100%, 97% across-sample, and 97% within-sample clustering approaches. A) The composition of unique sequence variants (i.e., 100% OTUs) in each sample, sorted by relative abundance, and filled by unique colors corresponding to unique sequence variants. The dominant OTUs within each sample assigned by B) 97% across-sample clustering and C) 97% within-sample clustering are shown by rectangles below each bar, with fill colors matching the unique sequence variants presented in (A) to indicate the representative sequence of the assigned OTU. Colored rectangles that span multiple bars indicate that the corresponding samples were assigned the same dominant OTU. Rectangles are annotated with the taxonomic assignment of the OTU (see Methods)."}
set.seed(42432) 
par(mfrow=c(1,1))
otubarplot2(samples=c("64","72","29","45","73","59","47","26","48","66","63"))
```

## *Symbiodinium* community composition

@@

```{r Fig3, fig.height=6, fig.width=4, fig.cap="Figure 3. *Symbiodinium* community composition of each sample. Samples are plotted as horizontal bars, sorted by species and shore of collection (north vs. south). Individual segments of each bar represent 97% within-sample OTUs, colored by *Symbiodinium* clade identity."}
#png(file="crap.png", width = 4, height=6, units="in", res=300)
par(mfrow=c(1,1), mar=c(2, 1.5, 2, 5), lwd=0.1, cex=1)
# Plot composition of 97% within-sample OTUs colored by clade
composition(phy97bs.f.p, col=taxcolors[factor(data.frame(tax_table(phy97bs.f.p))[order(data.frame(tax_table(phy97bs.f.p))$Subtype), ]$Clade, levels=c("A","B","C","D","F","G"))], legend=T)
#dev.off()
```

## Differences in *Symbiodinium* between shores

@@

```{r Table2, cache=T } 
set.seed(43789)  
shorestats97bs <- perms(phy97bs.f.p, group="Species", trt="Shore")
shorestats97bs$Species <- str_replace_all(
  shorestats97bs$Species,
  c("alcicornis"="*Millepora alcicornis*",
    "annularis"="*Orbicella annularis*",
    "astreoides"="*Porites astreoides*",
    "cavernosa"="*Montastraea cavernosa*",
    "cylindrus"="*Dendrogyra cylindrus*",
    "fragum"="*Favia fragum*",
    "furcata"="*Porites furcata*",
    "radians"="*Siderastrea radians*",
    "siderea"="*Siderastrea siderea*",
    "strigosa"="*Pseudodiplora strigosa*")
)
knitr::kable(shorestats97bs, digits=3, row.names=F, 
             caption="Mean overall, within-shore ('within'), and between-shore ('between') Bray-Curtis dissimilarities of the *Symbiodinium* communities in each host species, and PERMANOVA tests (partial R^2^ and p-values) for a difference between shores. Between-shore tests were not possible for *O. annularis* and *P. astreoides* since they were only sampled from one shore.")

@@

## Network analysis of *Symbiodinium* metacommunity
```{r networks,  cache=F, include=FALSE}
# Create dominant symbionts network
set.seed(12374)
domnet <- sppnet(phy=phy97bs.f.p, plot=F,
                 fun=function(x) length(x[x>0.5])/length(x), 
                 layout=layout.fruchterman.reingold)
# Calculate number of dominant symbionts connected to each species
dom.Nedges <- sapply(levels(data.frame(sample_data(phy97bs.f.p))$Species),
                     FUN=function(x) length(E(domnet)[from(x)]))
dom.Nedges <- sort(dom.Nedges, decreasing=T)

@@

```{r Fig4, fig.width=5, fig.height=5, fig.cap="Figure 4. Network analysis of abundant *Symbiodinium* taxa in each coral species. *Symbiodinium* OTUs (circular nodes) are connected to each coral species (square nodes) in which they ever occurred at >1% relative abundance within an individual; thickness of edges (i.e., links between coral and symbiont nodes) is relative to the proportion of individuals within the coral species in which the *Symbiodinium* OTU was present at >1%. Symbiont nodes are colored by clade identity, and sized proportionally to the number of coral species in which they were found."}

@@

```{r Fig5, fig.width=7.20472, fig.height=3.60236, fig.cap="Figure 5. Network analysis of dominant (A) and background (B) *Symbiodinium* taxa in each coral species (see Materials and Methods for details on network construction). Symbiont nodes are colored by clade identity, sized proportionally to the number of coral species to which they are connected."}

@@

## Symbiotic flexibility of host species

@@

```{r Fig6, cache=F, fig.width=3.5, fig.height=3.5, fig.cap="Figure 6. Symbiotic flexibility in coral species quantified as multivariate dispersion of *Symbiodinium* community composition (mean distance to centroid) $\\pm$ SE. Host taxa that do not share a letter are significantly different (p < 0.05)."}
betad97bs <- betad(phy97bs.f.t, group="Species")

@@

# Discussion

## Bioinformatic analysis of ITS2 diversity

@@

## *Symbiodinium* metacommunity ecology in St. John

@@

  In addition to hosting apparent generalist ITS2 lineages, most coral species sampled in St. John (except *M. cavernosa* and *P. astreoides*) hosted abundant *Symbiodinium* OTUs that were not abundant in any other corals, and are thus apparently more host-specific. Unique OTUs in the B1-radiation were abundant in *M. alcicornis* and *D. cylindrus*, while others from both the B1- and B19-radiations [see @LaJeunesse:2004p122] were abundant in *F. fragum*, *S. radians*, and *P. strigosa*. Several clade C OTUs similarly were only abundant in one species: C1144 in *P. strigosa*, C7 in *O. annularis*, C1 in *S. siderea*, C1c/C45 and C45a in *P. furcata*, and C46 in *S. radians*. Interestingly, *S. radians* was the only species in which none of the abundant symbionts occurred in any other host, which may reflect the unique ecology of *S. radians* as the only study species that typically forms small, encrusting colonies as adults.

@@

## Conclusions

@@

# Acknowledgements

@@

# References

### 2017-03-26 supp/STJ2012_analysis.Rmd
submission

```{r setup_2, include=FALSE}
# Load all R scripts in "R/" directory
source("R/functions.R", .GlobalEnv)
# Load package libraries
library(phyloseq); library(vegan); library(multcompView); library(reshape2); library(igraph); library(stringr)
# Set colors for plotting clades
taxcolors <- matrix(c("#8dd3c7", "#ffffb3", "#bebada", "#fb8072", "#80b1d3", "#fdb462"), 
                    dimnames=list(c("CladeA", "CladeB", "CladeC", "CladeD", "CladeF", "CladeG")))
```

@@

# Data pre-processing
### Import dataset
The [phyloseq](https://joey711.github.io/phyloseq/) package is used here to combine the OTU table, taxonomic assignments, and sample metadata into a single R data object (class 'phyloseq') to facilitate downstream analyses.
```{r import_taxa_data, cache=T}
# Import taxonomic assignment data from blast and nw
read.nw <- function(file) {
  nw <- read.table(file, stringsAsFactors=FALSE)
  nw <- nw[order(nw$otu),]
  return(nw)
}

@@

cladeAbund$Notus <- table(data.frame(tax_table(phy97bs.f.p))$Clade)
```

@@

##### Results:
- __*P. furcata* is the only species whose symbiont communities are significantly different between the north and south shores__, at both the 97% within-sample OTU and 100% OTU scales.

# Main findings
1. Symbiont communities are generally consistent with previous work on these species in the Caribbean, but high-throughput sequencing provides a much **more complete picture** of these communities’ composition.
2. South shore *P. furcata* are more likely to have clade A instead of C, and *S. siderea* more likely to have clade D instead of C.
3. **Within-sample OTU clustering** is more appropriate than clustering across samples.
4. We can apply **new types of analysis** with this much data, including network analysis and statistical tests on beta diversity.

>>>>

```{r setup_2, include=FALSE}
# Load custom R functions
source("R/functions.R", .GlobalEnv)
# Load package libraries
library(phyloseq); library(vegan); library(multcompView); library(reshape2); library(igraph); library(stringr)
# Set colors for plotting clades
taxcolors <- matrix(c("#8dd3c7", "#ffffb3", "#bebada", "#fb8072", "#80b1d3", "#fdb462"), 
                    dimnames=list(c("CladeA", "CladeB", "CladeC", "CladeD", "CladeF", "CladeG")))
```

@@

# Data pre-processing
### Import dataset
The [phyloseq](https://joey711.github.io/phyloseq/) package is used here to combine the OTU table, taxonomic assignments, and sample metadata into a single R data object (class 'phyloseq') to facilitate downstream analyses.
```{r import_taxa_data, cache=T}
# Import taxonomic assignment data from nw
read.nw <- function(file) {
  nw <- read.table(file, stringsAsFactors=FALSE)
  nw <- nw[order(nw$otu),]
  return(nw)
}

@@

cladeAbund$Notus <- table(data.frame(tax_table(phy97bs.f.p))$Clade)

@@
